---
title: "Bottom Trawl Survey Evaluation Report: `r params$stock_name`"
author: "`r params$author_name`"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    self-contained: true
params:
  author_name: "Liz Brooks"
  stock_name: "Plaid Fish"
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(here)
library(htmltools)

# Define the directory where plots are saved
plot_dir <- here("plots")

# Helper function to check if files exist for a specific unit
has_unit_files <- function(pattern_str) {
  length(list.files(plot_dir, pattern = pattern_str, full.names = TRUE)) > 0
}
```

## Survey Area Metrics

### Stratum Area Proportion
This plot illustrates the proportion of total stock area that each individual stratum represents for `r params$stock_name`.

```{r area-prop}
area_plot <- list.files(plot_dir, pattern = "stratum.area.proportion.plot", full.names = TRUE)

if (length(area_plot) > 0) {
  knitr::include_graphics(area_plot)
} else {
  cat("Stratum area proportion plot not found in plots directory.")
}
```

### Stock Area Sampled Tracking
Tracks the total proportion of the stock area successfully sampled across historical survey years. The dashed threshold sits at 95% coverage.

```{r area-sampled}
sampled_plot <- list.files(plot_dir, pattern = "stock.area.sampled.plot", full.names = TRUE)

if (length(sampled_plot) > 0) {
  knitr::include_graphics(sampled_plot)
} else {
  cat("Stock area sampled tracking plot not found in plots directory.")
}
```

---

## Stratum Mean Catch Trends

### Incomplete Sampling Discrepancies
Shows the calculated divergence in the stratified mean resulting strictly from years with incomplete spatial survey coverage. Use tabs below to toggle evaluation metrics.

::: {.panel-tabset}

```{r discrepancies-tabset}
#| output: asis

avail_discrep_units <- c()
if (has_unit_files("mean.incomplete.diff.plot\\.N\\.")) avail_discrep_units <- c(avail_discrep_units, "N")
if (has_unit_files("mean.incomplete.diff.plot\\.Kg\\.")) avail_discrep_units <- c(avail_discrep_units, "Kg")

if (length(avail_discrep_units) == 0 && has_unit_files("mean.incomplete.diff.plot")) {
  avail_discrep_units <- "Default"
}

for (u in avail_discrep_units) {
  tab_title <- switch(u, 
                      "N" = "Abundance (N/tow)", 
                      "Kg" = "Biomass (Kg/tow)", 
                      "Default" = "Survey Discrepancies")
  
  cat(paste0("#### ", tab_title, "\n\n"))
  
  diff_pattern <- if(u == "Default") "mean.incomplete.diff.plot" else paste0("mean.incomplete.diff.plot\\.", u)
  diff_plot <- list.files(plot_dir, pattern = diff_pattern, full.names = TRUE)
  if (length(diff_plot) > 0) {
    cat(paste0("![](", diff_plot, ")\n\n"))
  } else {
    cat("No discrepancy plot generated for this unit (this is normal if sampling was 100% complete across all years).\n\n")
  }
}
```

:::

### Stratum Mean Time-Series
Time series of strata means, by season. Use tabs below to toggle evaluation metrics.

::: {.panel-tabset}

```{r stratum-means-ts-tabset}
#| output: asis

avail_mean_units <- c()
if (has_unit_files("stratum.means.plot.page.*\\.N\\.")) avail_mean_units <- c(avail_mean_units, "N")
if (has_unit_files("stratum.means.plot.page.*\\.Kg\\.")) avail_mean_units <- c(avail_mean_units, "Kg")

if (length(avail_mean_units) == 0 && has_unit_files("stratum.means.plot.page")) {
  avail_mean_units <- "Default"
}

for (u in avail_mean_units) {
  tab_title <- switch(u, 
                      "N" = "Abundance (N/tow)", 
                      "Kg" = "Biomass (Kg/tow)", 
                      "Default" = "Survey Means")
  
  cat(paste0("#### ", tab_title, "\n\n"))
  
  mean_pattern <- if(u == "Default") "stratum.means.plot.page" else paste0("stratum.means.plot.page.*\\.", u)
  mean_pages <- list.files(plot_dir, pattern = mean_pattern, full.names = TRUE)
  
  if (length(mean_pages) > 0) {
    for (page in sort(mean_pages)) {
      page_num <- gsub(".*page\\.(\\d+)\\..*", "\\1", page)
      cat(paste0("##### Page ", page_num, "\n\n"))
      cat(paste0("![](", page, ")\n\n"))
    }
  } else {
    cat("No paginated stratum mean plots found for this unit.\n\n")
  }
}
```

:::

---

## Stratum Catch Contributions to Mean

This section breaks down how much individual strata contribute to the total calculated mean index across seasons. Use the main tabs to switch metrics, and the inner tabs to toggle between seasonal configurations or see normalized proportions.

::: {.panel-tabset}

```{r stratum-contributions-tabset}
#| output: asis

contrib_units <- c()
if (has_unit_files("str.mean.by.contr\\.N\\.")) contrib_units <- c(contrib_units, "N")
if (has_unit_files("str.mean.by.contr\\.Kg\\.")) contrib_units <- c(contrib_units, "Kg")

if (length(contrib_units) == 0 && has_unit_files("str.mean.by.contr")) {
  contrib_units <- "Default"
}

for (u in contrib_units) {
  tab_title <- switch(u, "N" = "Abundance (N)", "Kg" = "Biomass (Kg)", "Default" = "Survey Contribution")
  cat(paste0("### ", tab_title, "\n\n"))
  
  p_comb   <- if(u == "Default") "^str.mean.by.contr\\." else paste0("^str.mean.by.contr\\.", u)
  p_fall   <- if(u == "Default") "str.mean.FALL.by.contr" else paste0("str.mean.FALL.by.contr\\.", u)
  p_spring <- if(u == "Default") "str.mean.SPRING.by.contr" else paste0("str.mean.SPRING.by.contr\\.", u)
  
  p_comb_prop   <- if(u == "Default") "str.mean.by.contr.prop\\." else paste0("str.mean.by.contr.prop\\.", u)
  p_fall_prop   <- if(u == "Default") "str.mean.FALL.by.contr.prop" else paste0("str.mean.FALL.by.contr.prop\\.", u)
  p_spring_prop <- if(u == "Default") "str.mean.SPRING.by.contr.prop" else paste0("str.mean.SPRING.by.contr.prop\\.", u)
  
  f_comb   <- list.files(plot_dir, pattern = p_comb, full.names = TRUE)
  f_fall   <- list.files(plot_dir, pattern = p_fall, full.names = TRUE)
  f_spring <- list.files(plot_dir, pattern = p_spring, full.names = TRUE)
  
  f_comb_prop   <- list.files(plot_dir, pattern = p_comb_prop, full.names = TRUE)
  f_fall_prop   <- list.files(plot_dir, pattern = p_fall_prop, full.names = TRUE)
  f_spring_prop <- list.files(plot_dir, pattern = p_spring_prop, full.names = TRUE)
  
  cat("::::: {.panel-tabset}\n\n")
  
  if(length(f_comb) > 0) {
    cat("#### Combined Seasons (Absolute)\n\n")
    cat(paste0("![](", f_comb, ")\n\n"))
  }
  if(length(f_comb_prop) > 0) {
    cat("#### Combined Seasons (Normalized Proportions)\n\n")
    cat(paste0("![](", f_comb_prop, ")\n\n"))
  }
  if(length(f_fall) > 0) {
    cat("#### Fall Survey (Absolute)\n\n")
    cat(paste0("![](", f_fall, ")\n\n"))
  }
  if(length(f_fall_prop) > 0) {
    cat("#### Fall Survey (Normalized Proportions)\n\n")
    cat(paste0("![](", f_fall_prop, ")\n\n"))
  }
  if(length(f_spring) > 0) {
    cat("#### Spring Survey (Absolute)\n\n")
    cat(paste0("![](", f_spring, ")\n\n"))
  }
  if(length(f_spring_prop) > 0) {
    cat("#### Spring Survey (Normalized Proportions)\n\n")
    cat(paste0("![](", f_spring_prop, ")\n\n"))
  }
  
  cat(":::::\n\n")
}
```

:::

---

## Effort Tracking (Number of Tows)

### Annual Tow Totals & CV Matrix

::: {.panel-tabset}
### Total Tows per Season
```{r total-tows}
tot_tows <- list.files(plot_dir, pattern = "ntows.yr.season.plot", full.names = TRUE)
if (length(tot_tows) > 0) knitr::include_graphics(tot_tows)
```

### CV Variations over Time
```{r cv-trends-tabset}
#| output: asis

cv_units <- c()
if (has_unit_files("cv.yr.season.plot\\.N\\.")) cv_units <- c(cv_units, "N")
if (has_unit_files("cv.yr.season.plot\\.Kg\\.")) cv_units <- c(cv_units, "Kg")
if (length(cv_units) == 0 && has_unit_files("cv.yr.season.plot")) cv_units <- "Default"

if (length(cv_units) > 1) cat("::::: {.panel-tabset}\n\n")

for (u in cv_units) {
  if (length(cv_units) > 1) {
    t_title <- switch(u, "N" = "Abundance (N) CV", "Kg" = "Biomass (Kg) CV")
    cat(paste0("#### ", t_title, "\n\n"))
  }
  cv_pat <- if(u == "Default") "cv.yr.season.plot" else paste0("cv.yr.season.plot\\.", u)
  cv_f <- list.files(plot_dir, pattern = cv_pat, full.names = TRUE)
  if (length(cv_f) > 0) cat(paste0("![](", cv_f, ")\n\n"))
}

if (length(cv_units) > 1) cat(":::::\n\n")
```
:::

### Detailed Tow Distributions by Stratum
::: {.panel-tabset}

### Fall (All Years)
```{r tows-fall-all}
#| output: asis
fall_all <- list.files(plot_dir, pattern = "ntow.fall.by.stratum.plot.page", full.names = TRUE)
for (f in sort(fall_all)) { cat(paste0("![](", f, ")\n\n")) }
```

### Fall (Bigelow Years)
```{r tows-fall-bigelow}
#| output: asis
fall_big <- list.files(plot_dir, pattern = "ntow.fall.by.stratum.bigyrs.plot.page", full.names = TRUE)
for (f in sort(fall_big)) { cat(paste0("![](", f, ")\n\n")) }
```

### Spring (All Years)
```{r tows-spring-all}
#| output: asis
spring_all <- list.files(plot_dir, pattern = "ntow.spring.by.stratum.plot.page", full.names = TRUE)
for (s in sort(spring_all)) { cat(paste0("![](", s, ")\n\n")) }
```

### Spring (Bigelow Years)
```{r tows-spring-bigelow}
#| output: asis
spring_big <- list.files(plot_dir, pattern = "ntow.spring.by.stratum.bigyrs.plot.page", full.names = TRUE)
for (s in sort(spring_big)) { cat(paste0("![](", s, ")\n\n")) }
```

:::
