---
title: "MSE Setup"
author: "Jonathan J. Deroba"
date: "`r Sys.Date()`"
output:
  html_document:
    code_fold: hide
link-citations: yes
csl: "canadian-journal-of-fisheries-and-aquatic-sciences.csl"
bibliography: MSE.bib
urlcolor: blue  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Setup

Following the MSE vignette from Chengxue Li at https://lichengxue.github.io/whamMSE/

```{r}
library(wham)
library(whamMSE)
library(fmsb)
library(ggpubr)

main.dir = here::here()
```

## Modify for short-lived

Basic info for a short-lived species. Based on @wiedenmann_evaluation_2017, Table 3 B and C.

A short-lived species recruits at 1 year, lives to 7, mean M is 0.4, 50% mature at age 2 (1.75 in Wiedenmann). 

Modify from https://lichengxue.github.io/whamMSE/03.Management-Strategy-Evaluation.html

Start with a 30 year burn in and 30 MSE years. First try constant low F. Then work up to complex case.

Historical fishing pattern is "F-H-L".

Setup the ideal baseline with assessments 2 years, both surveys, and "good" data.

Now set up as in new Single Stock MSE vignette: https://lichengxue.github.io/whamMSE/11.Single-Stock-MSE.html


```{r}

year_start  <- 1  # starting year in the burn-in period
year_end    <- 30  # end year in the burn-in period
MSE_years   <- 10     # number of years in the feedback loop
# Note: no need to include MSE_years in simulation-estimation 

info <- generate_basic_info(n_stocks = 1, 
                            n_regions = 1,
                            n_indices = 2,
                            n_fleets = 1,
                            n_seasons = 1,
                            base.years = year_start:year_end,
                            n_feedback_years = MSE_years,
                            life_history = "short",
                            n_ages = 7,
                            F_info = list(F.year1 = 0.33, Fhist = "F-H-L", Fmax = 2.5, Fmin = 1, change_time = 0.5, user_F = NULL), #0.33 is approximately F40% based on Wiedenmann
                            catch_info = list(catch_cv = 0.1, catch_Neff = 100),
                            index_info = list(index_cv = c(0.25,0.3), index_Neff = c(100,100), fracyr_indices = c(0.25,0.75), q = c(0.000001,0.000001),units_indices=2,units_index_paa=2),
                            fracyr_spawn = 0.85) 

basic_info = info$basic_info # collect basic information
catch_info = info$catch_info # collect fleet catch information
index_info = info$index_info # collect survey information
F_info = info$F # collect fishing information

# see more details using ?generate_basic_info
```

Movement, do we need this? 1 stock 1 region model, turn it off

```{r}
basic_info <- generate_NAA_where(basic_info = basic_info, move.type = NULL) # "bidirectional" movement

move <- NULL

#move <- generate_move(basic_info = basic_info, move.type = 2, move.rate = c(0.3), #,0.1
#                     move.re = "iid_y", move.sigma = 0.5, prior.sigma = 0.5, use.prior = TRUE)
```
Adjust for 1 area 1 stock, short lived M=0.4

```{r}
n_stocks  <- as.integer(basic_info['n_stocks'])
n_regions <- as.integer(basic_info['n_regions'])
n_fleets  <- as.integer(basic_info['n_fleets'])
n_indices <- as.integer(basic_info['n_indices'])
n_ages    <- as.integer(basic_info['n_ages'])

# Selectivity Configuration
fleet_pars <- c(3,1)
index_pars1 = c(4,1)
index_pars2 = c(3,0.5)
sel <- list(model=rep("logistic",n_fleets+n_indices),
            initial_pars=c(rep(list(fleet_pars),n_fleets),list(index_pars1, index_pars2)))
# M Configuration
M <- list(model="constant",initial_means=array(0.4, dim = c(n_stocks,n_regions,n_ages)))
```



```{r}
sigma      <- "rec+1"
re_cor     <- "iid"
ini.opt    <- "equilibrium" # option   <- c("age-specific-fe", "equilibrium")

# Set para. mean rec
#alpha <- 12
#beta  <- 1.5e-4
meanrec=2500000  #1*10^9

# Set sigma for NAA
NAA_sig <- 0.26
sigma_vals = array(NAA_sig, dim = c(n_stocks, n_regions, n_ages)) # n_stocks x n_regions x n_ages"
rec_sig=0.8
sigma_vals[[1]]=rec_sig

# Set initial NAA for each stock
# log_N1  <- rep(log(meanrec/2), n_stocks) # Create difference between stocks
N1_pars <- generate_ini_N1(basic_info, ini.opt, log_N1 = log(meanrec/2), log_N1_F = 0.5)

NAA_re <- list(N1_model=rep(ini.opt,n_stocks),
               sigma=rep(sigma,n_stocks),
               cor=rep(re_cor,n_stocks),
               recruit_model = 2,
               recruit_pars = rep(list(c(meanrec)),n_stocks), # assume same constant s-r 
               sigma_vals = sigma_vals,
               N1_pars = N1_pars)#,
               #NAA_where = basic_info$NAA_where)

# recruit_model = 1: estimating annual recruitments as fixed effects or a random walk if NAA_re$sigma specified
# recruit_model = 2: estimating a mean recruitment with annual recruitments as random effects
# recruit_model = 3: Beverton-Holt stock-recruitment with annual recruitments as random effects
# recruit_model = 4: Ricker stock-recruitment with annual recruitments as random effects

# 1. recruit_pars: a list (length = n_stocks) of vectors of initial parameters for recruitment model. 
# If $recruit_model is 3 (B-H) or 4 (Ricker), parameters are "alpha" and "beta".

# 2. sigma_vals: Initial standard deviation values to use for the NAA deviations. Values are not used if recruit_model = 1 
# If sigma="rec": must be a list (length = n_stocks) of single values
# If sigma="rec+1": a list (length = n_stocks) of 2 values must be specified. First is for the first age class (recruits), second is for all other ages.
```

Make inputs for the OM

```{r}
 input <- prepare_wham_input(basic_info = basic_info, 
                            selectivity = sel, 
                            M = M, 
                            NAA_re = NAA_re, 
                            move = move,
                            catch_info = catch_info, 
                            index_info = index_info, 
                            F = F_info)

# IMPORTANT!#
# This appears to be due to the initial F value for the Newton iterations to find F40. The default value is too high for long-lived fish
input$data$FXSPR_init[] <-0.3 # 0.01 # change to a low value

```

Make the OM

```{r}
random = input$random # check what processes are random effects
input$random = NULL # so inner optimization won't change simulated RE
om <- fit_wham(input, do.fit = F, do.brps = T, MakeADFun.silent = TRUE)
# Note: do.fit must be FALSE (no modeling fitting yet)
```


One dataset

```{r}
om_with_data <- update_om_fn(om, seed = 123, random = random)
```

Assessment specifications, every 2 years as a start, MP could be annual

```{r}
assess.interval <- 2 # 
base.years      <- year_start:year_end # Burn-in period
first.year      <- head(base.years,1)
terminal.year   <- tail(base.years,1)
assess.years    <- seq(terminal.year, tail(om$years,1)-assess.interval,by = assess.interval)

mods <- list() # Create a list to save MSE outputs
```

Try a simple EM, looks like EM1 [here](https://lichengxue.github.io/whamMSE/03.Management-Strategy-Evaluation.html#10_Description_of_assessment_models:)

```{r}
# this matches the OM
# n_stocks = n_regions = n_fleets = n_indices = 1
# 
# sel_em <- list(model=rep("logistic",n_fleets+n_indices),
#                initial_pars=c(rep(list(fleet_pars),n_fleets),rep(list(index_pars),n_indices)))
# 
NAA_re_em <- list(N1_model="equilibrium",sigma="rec+1",cor="iid")
# 
# M_em <- list(model="constant",initial_means=array(0.1, dim = c(n_stocks,n_regions,n_ages))) # this is misspecified M

mods[[1]] = loop_through_fn(om = om_with_data,
                            em_info = info, 
                            random = random,
                            M_em = M, # use OM M
                            sel_em = sel, # use OM sel
                            NAA_re_em = NAA_re_em, # use rec assumed random around the mean instead, help runtime (est B-H is difficult)
                            move_em = NULL,
                            age_comp_em = "multinomial",
                            # Here is the correct code: separate.em = FALSE also works for one-area model
                            em.opt = list(separate.em = FALSE, separate.em.type = 1, 
                                          do.move = FALSE, est.move = FALSE),
                            assess_years = assess.years, 
                            assess_interval = assess.interval, 
                            base_years = base.years,
                            year.use = 20, # number of years of data you want to use in the assessment model
                            add.years = TRUE, # extends assessment time series instead of moving window of year.use years
                            seed = 123,
                            save.sdrep = FALSE, 
                            save.last.em = TRUE,
                            FXSPR_init = 0.3) # IMPORTANT!

###try plotting this one
#mods = list()

#for (nsim in 1:10) {
 # file.names <- file.path(sub.dir,paste0("Mod",1:6,sprintf("_%03d.RDS",nsim))) # path where you saved the results from Ex5.
#  mods[[nsim]] <- lapply(file.names, readRDS)
#}

#plot_mse_output(mods = mods, main.dir = getwd(), dpi = 150,
 #               use.n.years = 5, # The last 10 years will be used for comparison of long-term performance 
  #              base.mod = 1, # "correct" EM that can be used as a reference
   #             short.term = NULL, # if NULL, results of first 2 years will be used for comparison of short-term performance
    #            out.type = "html") # Options: png, pdf, html 

```

Try same model with poorer survey in the future

See https://lichengxue.github.io/whamMSE/15.Reduce-Data-Availability-In-Assessment.html#6_Increase_observation_error_and_reduce_sample_size_for_survey_indices_in_the_feedback_period 

```{r}
agg_index_sigma = input$data$agg_index_sigma
agg_index_sigma[21:40,] = 1 # Increase CV for both survey indices in the feedback period
index_Neff = input$data$index_Neff
index_Neff[21:40,] = 30 # Decrease ESS for both survey indices in the feedback period

input <- update_input_index_info(input, agg_index_sigma, index_Neff) # Update input file
```

Generate new om and dataset, keep same assessment interval as above

```{r}
om <- fit_wham(input, do.fit = F, do.brps = T, MakeADFun.silent = TRUE)
om_with_data <- update_om_fn(om, seed = 123, random = random)
```

Run new loop

```{r}
mods[[2]] = loop_through_fn(om = om_with_data,
                            em_info = info, 
                            random = random,
                            M_em = M, # use OM M
                            sel_em = sel, # use OM sel
                            NAA_re_em = NAA_re_em, # use rec assumed random around the mean instead, help runtime (est B-H is difficult)
                            move_em = NULL,
                            age_comp_em = "multinomial",
                            # Here is the correct code: separate.em = FALSE also works for one-area model
                            em.opt = list(separate.em = FALSE, separate.em.type = 1, 
                                          do.move = FALSE, est.move = FALSE),
                            assess_years = assess.years, 
                            assess_interval = assess.interval, 
                            base_years = base.years,
                            year.use = 20, # number of years of data you want to use in the assessment model
                            add.years = TRUE,
                            update_index_info  = list(agg_index_sigma = agg_index_sigma, index_Neff = index_Neff), # Must have this!
                            seed = 123,
                            save.sdrep = FALSE, 
                            save.last.em = TRUE,
                            # Initial F should be much higher than 0.01 otherwise F brp can't be found 
                            FXSPR_init = 0.3) # IMPORTANT!

```


Visualize 2 mods (modified from https://lichengxue.github.io/whamMSE/03.Management-Strategy-Evaluation.html#12_Compare_model_performance)

This compares a good survey in the future (1) with a poor survey in the future (2)

```{r}

par(mfrow = c(1,2))
SSB_s1 <- lapply(mods, function(mod) mod$om$rep$SSB[,1])
plot(SSB_s1[[1]], type = "l", col = "blue", ylab = "SSB", xlab = "Year", ylim=c(0,2000000), main = "OM SSB", lwd = 2) 
lines(SSB_s1[[2]], col = "red",lty = 2, lwd = 2)
legend("topleft",legend = paste0("EM ", 1:2), col = c("blue","red"),lty=1:2, cex=0.8)

Catch_s1 <- lapply(mods, function(mod) mod$om$rep$pred_catch[,1])
plot(Catch_s1[[1]], type = "l", col = "blue", ylab = "Catch", xlab = "Year", ylim=c(0,1000000), main = "OM Catch", lwd = 2)
lines(Catch_s1[[2]], col = "red",lty = 2, lwd = 2)

```
Now add the one with no survey, from Cheng's code

```{r}
# ---------------------------------
# Actually the above section (CV and Neff) will be useless because we will not have survey data in the feedback period), see code below:
# ---------------------------------
remove_agg = TRUE # remove a aggregate index for some years
remove_agg_pointer = 1 # target on index 1
remove_agg_years = 21:40 # all feedback years
remove_paa = TRUE # Also remove age comp for that index 
remove_paa_pointer = 1 # target on index 1
remove_paa_years = 21:40 # all feedback years

input <- update_input_index_info(input, agg_index_sigma, index_Neff,
                                 remove_agg, remove_agg_pointer, remove_agg_years,
                                 remove_paa, remove_paa_pointer, remove_paa_years) # Update input file
```

Generate new om and dataset, keep same assessment interval as above

```{r}
om <- fit_wham(input, do.fit = F, do.brps = T, MakeADFun.silent = TRUE)
om_with_data <- update_om_fn(om, seed = 123, random = random)
```

Run new loop

```{r}
# this simulation actually represents the data-missing scenario.
mods[[3]] = loop_through_fn(om = om_with_data,
                            em_info = info, 
                            random = random,
                            M_em = M, # use OM M
                            sel_em = sel, # use OM sel
                            NAA_re_em = NAA_re_em, # use rec assumed random around the mean instead, help runtime (est B-H is difficult)
                            move_em = NULL,
                            age_comp_em = "multinomial",
                            # Here is the correct code: separate.em = FALSE also works for one-area model
                            em.opt = list(separate.em = FALSE, separate.em.type = 1, 
                                          do.move = FALSE, est.move = FALSE),
                            assess_years = assess.years, 
                            assess_interval = assess.interval, 
                            base_years = base.years,
                            year.use = 20, # number of years of data you want to use in the assessment model
                            add.years = TRUE,
                            # Must have the below code! Otherwise the EM will assume a constant CV and Neff (same as the CV and Neff in the first year) for the survey
                            update_index_info  = list(agg_index_sigma = agg_index_sigma, index_Neff = index_Neff), # Must have this!
                            seed = 123,
                            save.sdrep = FALSE, 
                            save.last.em = TRUE,
                            # Initial F should be much higher than 0.01 otherwise F brp can't be found 
                            FXSPR_init = 0.1) # IMPORTANT!

```

Visualize 3 mods (modified from https://lichengxue.github.io/whamMSE/03.Management-Strategy-Evaluation.html#12_Compare_model_performance)

```{r}

par(mfrow = c(1,2))
SSB_s1 <- lapply(mods, function(mod) mod$om$rep$SSB[,1])
plot(SSB_s1[[1]], type = "l", col = "blue", ylab = "SSB", xlab = "Year", ylim=c(2e4,3e6), main = "OM SSB", lwd = 2) 
lines(SSB_s1[[2]], col = "red",lty = 2, lwd = 2)
lines(SSB_s1[[3]], col = "orange",lty = 3, lwd = 2)

legend("topleft",legend = c("EM1_good_survey", "EM2_poor_survey","EM3_missing_survey"), col = c("blue","red","orange"),lty=1:3, cex=0.8)

Catch_s1 <- lapply(mods, function(mod) mod$om$rep$pred_catch[,1])
plot(Catch_s1[[1]], type = "l", col = "blue", ylab = "Catch", xlab = "Year", ylim=c(0,1.5e6), main = "OM Catch", lwd = 2)
lines(Catch_s1[[2]], col = "red",lty = 2, lwd = 2)
lines(Catch_s1[[3]], col = "orange",lty = 3, lwd = 2)

```


# References